Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-07-21 thesis/introduction.tex
introduction updates

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\),
e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
galaxies, dwarfs are the most numerous galaxies in the universe. Dwarf
galaxies are highly \emph{dark-matter dominated}, with mass to light
ratios sometimes exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of
their small total masses, many dwarf galaxies are \emph{quenched}, with
little to no recent star formation. Their stellar populations are
\emph{relics} from the early universe, consisting of many of the oldest
and most metal poor stars. Understanding the properties of dwarf
galaxies thus has relevance across astronomy, from cosmological
structure formation on the smallest scale to extremely metal-poor
stellar populations to chemical evolution to galactic dynamics. Of the
classical dwarfs, the Sculptor and Ursa Minor dwarf spheroidal galaxies
stand out with a more extended density profile relative to an
exponential, hinting at tidal effects. As a case study in tidal effects
and our understanding of dwarf galaxy formation and evolution, we aim to
understand the dynamical history of these galaxies.

In this section, we first describe the observational history of dwarf
galaxies, what \emph{Gaia} is and how this telescope has changed our
understanding of the local group, the present state of dwarf galaxy
knowledge, and why Sculptor and Ursa Minor stand out. We then describe
the role of dwarf galaxies in cosmology, their formation and our current
theory of hierarchichal structure formation. We later discuss idealized
simulations, tidal effects, break radii, and recent developments in
tidal simulations. Finally, we provide a roadmap to the thesis as a
whole at the end of this section.

@@

The Magellanic clouds are the most prominent of the Milky Way satellites
and were likely always known to humans in the southern hemisphere.
However, the first discovered dwarf galaxies in 1938, Sculptor and
Fornax, heralds i n an era of discovery and investigation into a novel
astronomical system. \citet{shapley1938} present Sculptor and Fornax as
some kind of \emph{stellar system} resembling the Magellanic clouds and
globular clusters. The first direct hints that these stellar systems are
galaxies originates from early spectroscopic work
\citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
\citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
\citet{olszewski1998}.]{aaronson1983}. Based on their velocity
dispersions given their physical sizes, the mass to light ratios of
these dwarf galaxies is far larger than expected for star clusters
without dark matter. In general, these objects have reached a consensus
that they have high velocity dispersion, old spheroidal stellar
populations, and low stellar masses. While we now know the basic
properties of dwarf galaxies, tremendous progress has been made in the
past 10-20 years with large area deep photometric surveys, multi-object
spectrographs, high resolution chemistry, and theoretical work,
unveiling that dwarf galaxies are each unique systems with different
formation pathways \citep[e.g.][]{simon2019}.

Dwarf galaxies span a large range of size and morphology.
Fig.~\ref{fig:galaxy_images} displays several images of dwarf galaxies.
The largest in the local group but not always considered a dwarf galaxy,
the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. From
this figure, we notice that while large LMC-like dwarf galaxies exhibit
complex structure such as a bar, young blue stars, and irregular shapes,
the classical and faint dwarf galaxies are composed mostly of very old,
faint, red stars, very diffuse and often large on the sky if local.
These galaxies are typically classified as dwarf spheroidals (dSph),
however we typically omit the dSph designation and refer to these
galaxies by their constellation hereafter.

@@

\textbf{Classical systems}, the brightest of the local group
(non-Magellanic) dwarf galaxies, are perhaps the systems best studied
and with the most stringingly derived properties. Defined\footnote{Note
  that some work like \citet{simon2019} instead define classical dwarfs
  based on luminosity as compared to ultra faint dwarfs. This may
  include galaxies such as Antlia I, Crater II, Canes Venatici I, and
  Sagittarius. We stick with the historical definition for simplicity
  and since the Gaia observations are most numerous for the historical
  classicals.} as the first six discovered dwarf spheroidal galaxies,
(Sculptor, Ursa Minor, Fornax, Draco, Leo I, Leo II, and Carina),
classical systems are typically old, spheroidal, and gas-poor. While
extending 1-2 degrees across the sky (e.g.
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
these dwarfs contain large numbers of bright (giant) stars allow for
thousands of stars to be observed with deep photometry and spectroscopy
\citep[e.g.][\citet{pace+2020}]{tolstoy+2023}. As a result, the
determination of fundamental properties such as the position, size,
orientation, distance \citep[from 100s of RRL stars,
e.g.][]{tran+2022, garofalo+2025}, proper motions (from \emph{Gaia},
e.g. \citet{battaglia+2022}, \citet{MV2020a}), line-of-sight (LOS)
velocity and dispersion \(\sigma_v\), are all relatively well
constrained today and show excellent convergence among different works.
However, ongoing research still is refining our understanding of the
detailed structure of Milky Way satellites, hinting that these objects
may be more complex than at first glance.

@@

For Local group dwarf spheroidals, \emph{Gaia} has allowed for many of
the first proper motion measurements. While the proper motion
uncertainty on a typical dwarf member star is often with large
uncertainties, by combining the proper motions for 100s to 1000s of
stars in \emph{Gaia}, incredibly precise proper motion measurements can
be determined \citep[e.g.][]{MV2020a}, sometimes only limited by
\emph{Gaia}'s systematic error floor. Proper motions have ushered in a
new dynamical era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's (short term) orbit under a given MW
potential.

@@

\subsubsection{Exponential stellar density
profiles}\label{exponential-stellar-density-profiles}

@@

Since \citet{faber+lin1983} demonstrated that an Exponential density
profile is a reasonable empirical fit to dwarf galaxies, many works have
fitted an Exponential profile (\citet{Eskridge} 1988a,c; Hodge et al
1991a,b; \citet{IH1995}, ) in addition to Sérsic, King and various other
profiles. \citet{lelli+2014} (example of using exponential.)

\citet{wang+2019} for fornax, \citet{mcconnachie+irwin2006} for Exp,
Plummer, King in other galaxies.

However, even starting from \citet{aparicio1997}, some people have noted
deviations from this simple empirical rule. Additionally,
\citet{herrmann+hunter+elmegreen2013},
\citet{herrmann+hunter+elmegreen2016} note that at least in a
photometric sample of more distant dwarf disky/Irr/blue compact dwarfs,
that many dwarfs show deviations from exponential profiles.

@@

\subsubsection{Hints of tides and
disruptions}\label{hints-of-tides-and-disruptions}

\subsubsection{Kinematics and structure}\label{kinematics-and-structure}

@@

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018) Sérsic fits, 2. Tran et al. (2022) RR lyrae distance, 3. Alan W. McConnachie and Venn (2020b), 4. Arroyo-Polonio et al. (2024). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & –\\
distance modulus & $19.60 \pm 0.05$ & 2\\
distance & $83.2 \pm 2$ kpc & –\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & –\\
$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
ellipticity & $0.37 \pm 0.01$ & –\\
position angle & $94\pm1^\circ$ & –\\
$M_V$ & $-10.82\pm0.14$ & –\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018) Sérsic fits, (2) Garofalo et al. (2025) RR lyrae distance, (3) Alan W. McConnachie and Venn (2020a), (4) Pace et al. (2020), average of MMT and Keck results. }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & –\\
distance modulus & $19.23 \pm 0.11$ & 2\\
distance & $70.1 \pm 3.6$ kpc & –\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & –\\
LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
$\sigma_v$ & $8.6 \pm 0.3$ & –\\
$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & –\\
position angle & $50 \pm 1^\circ$ & –\\
$M_V$ & $-9.03 \pm 0.05$ & –\\
\bottomrule
\end{tabular}
\end{table*}

\section{Cosmological and theoretical
context}\label{cosmological-and-theoretical-context}

The leading theory of cosmology, \(\Lambda\)CMD (cold dark matter) holds
that the universe is composed of about 68\% dark energy (\(\Lambda\)),
27\% dark matter (DM), and 5\% regular baryons
\citep{planckcollaboration+2020}. While baryonic matter is the only
matter we can directly understand or observe, we still know the general
properties of dark matter and dark energy. Dark energy causes the
acceleration of the expansion of the universe on large scales. We do not
discuss dark energy here as it does not substantially affect the local
group today. Dark matter makes up the vast majority of mass in galaxies.
Typically, galaxies have baryonic to dark matter ratios of between 1:5
to beyond 1:100. The faintest dwarf galaxies harbouring the largest
proportion of dark matter. In \(\Lambda\)CMD, dark matter is assumed to
only interact gravitationally with itself and other matter (transparent
to light, or \emph{dark}). Dark matter is also created \emph{cold},
i.e.~typical velocities much smaller than the speed of light in the
early universe. Dark matter is a key ingredient in cosmological
structural formation, dynamical galaxy structure, and galaxy-galaxy
interactions.

\subsection{Formation of dwarf
galaxies}\label{formation-of-dwarf-galaxies}

The very early universe was almost featureless. As observed from the
cosmic microwave background created at redshift \(z\sim 1100\), the
temperature is a near perfect blackbody with fluctuations only at 1 part
in 10,000. These tiny perturbations began to grow as the universe ages
and expands. Initially hot and dense, baryonic matter was prevented from
collapsing by thermal / radiation pressure. However, dark matter was
uninhibited, freely passing through the universe and collapsing into
denser regions. Each self-gravitating overdensity of dark matter is
known as a \emph{halo}. As the universe continued to expand and cool
down, baryonic matter began to collapse into dark matter halos.
Eventually, the densest regions of baryons situated near the centres of
halos would become the first stars of the first galaxies.

Halos rarely evolve as isolated systems. Instead, structure formation is
\emph{hierarchical}, where smaller \emph{subhalos} are accreted into
larger halos which may accrete into yet larger halos. The most massive
halos are made up of thousands of subhalos (verify), with each subhalo
becoming entire galaxies making up a galaxy cluster. But, even the
smallest galaxies reside in a \emph{halo}, with the dwarf galaxies
representing entities with simpler \emph{assembly histories}.

From cosmological simulations, we know that across many orders of
magnitude, dark matter halos appear self-similar. In
\citet{NFW1996};\citet{NFW1997}, hereafter NFW, the authors observe that
the radial density profiles \(\rho(r)\) follow a two-parameter law:
\begin{equation}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2}
}\end{equation} where \(r_s\) and \(\rho_s\) are the scale radius and
density (see below). This density profile has shown remarkable success
as a description of \(\Lambda\)CMD halos \citep[REFS, but see
also][darkEXP, ect.]{navarro2010}. This density profile is \emph{cuspy},
where the density continuously rises like \(\rho \sim 1/r\) at the
centre. Also note that the total mass of the NFW profile logarithmically
diverges. Typically, the virial mass of NFW halos is discussed instead,
defined as \(M_{200}\), the mass within a region \(r_{200}\) containing
a mean density 200 times the critical density of the universe
\(\rho_{\rm crit}\), so
\(M_{200} = M(r<r_{200}) = 200\ (4\pi/3) \ r_{200}^3\ \rho_{\rm crit}\).

There are several parameterizations of the NFW profile. \citet{NFW1996}
initially write the halo in terms of \(r_s\) and a concentration
\(c= r_{200} / r_s\) relating the virial radius \(r_{200}\) to \(r_s\).
. In this case,
\(\rho_s = \rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\). However, we
chose to discuss halos in terms of velocity space. The circular velocity
of the NFW halo, \(v_{\rm circ}(r) = \sqrt{G M(r) / r}\) related to the
enclosed mass \(M(r)\), reaches a maximum of \(v_{\rm max}\) at
\(r_{\rm max} = \alpha\,r_s\) where \(\alpha\approx2.16258\). Thus
\(r_{\rm max}\) is directly related to the scale radius and
\(v_{\rm max}\) is related to the total mass.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\),
e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
galaxies, dwarfs are the most numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios sometimes
exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
masses, many dwarf galaxies are \emph{quenched}, with little to no
recent star formation. Their stellar populations are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
the smallest scale to extremely metal-poor stellar populations to
chemical evolution to galactic dynamics. Of the classical dwarfs, the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

In this section, we first describe cosmological structure formation and
the role of dwarf galaxies. Then, we explore the observational history
of dwarf galaxies, the \emph{Gaia} telescope, open questions, and why
Sculptor and Ursa Minor stand out. We discuss idealized simulations,
tidal effects, break radii, and recent developments in tidal
simulations. Finally, we provide a roadmap to the thesis as a whole at
the end of this section.

\section{Cosmological context}\label{cosmological-context}

We only understand a tiny fraction of the composition of the universe.
The leading theory of cosmology, \(\Lambda\)CDM (cold dark matter),
posits that the universe is composed of about 68\% dark energy
(\(\Lambda\)), 27\% dark matter (DM), and 5\% regular baryons\footnote{Astronomers
  like to change definitions of words. \emph{Baryons} here means
  baryons+leptons, i.e.~any standard model massive fermion. The photon
  energy density is negligible today} \citep{planckcollaboration+2020}.
While the composition of dark matter and dark energy remains elusive, we
know their general properties. Dark energy causes the acceleration of
the expansion of the universe on large scales. We do not discuss dark
energy here---it does not substantially affect the local group today.
Dark matter, instead, makes up the vast majority of mass in galaxies.
Typically, galaxies have baryonic to dark matter ratios of between 1:5
to beyond 1:1000 for faint dwarf galaxies. In \(\Lambda\)CMD, dark
matter is assumed to only interact gravitationally, passing through
matter without effect (transparent to light, or \emph{dark}). Dark
matter is also \emph{cold}, i.e.~typical velocities much smaller than
the speed of light in the early universe. Implications of dark matter
properties range from cosmological structural formation, galaxy
structure, and galaxy interactions.

\subsection{Structure formation and dwarf
galaxies}\label{structure-formation-and-dwarf-galaxies}

The very early universe was almost featureless. Our earliest
observations of the universe stem from the cosmic microwave background
(CMB)---revealing a uniform, isotropic, near-perfect blackbody emission.
But tiny perturbations in the CMB, temperature fluctuations of 1 part in
10,000, became the seeds of large-scale cosmological structure. Governed
by cosmological expansion, gravitational collapse, and baryonic physics,
each perturbation grows into larger structures. Initially, baryonic
matter was coupled to radiation and resisted collapse. Dark matter, only
influenced by gravity instead, freely collapsed into the first
structures. Each self-gravitating overdensity of dark matter is known as
a \emph{halo}. After recombination, where electrons combined with atomic
nuclei to form atoms, baryons decoupled from radiation and fell into the
dark matter halos. The densest pockets of baryons later formed the first
stars and galaxies.

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos \citep[e.g.][\citet{blumenthal+1986},
\citet{white+rees1978}, \citet{white+frenk1991},
\citet{somerville+dave2015}]{blumenthal+1984}. Structure formation
happens on a wide range of scales. The largest structures become the
cosmic web---composed of voids, filaments, and clusters---and the
smallest structures directly observable host dwarf galaxies.
Hierarchical assembly is evident through the large scale structure of
the universe, remnants of past mergers within the milky way, and tidal
disruption of dwarf galaxies and their streams around nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy structure, formation, and tidal evolution.
The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

In \(\Lambda\)CDM cosmological simulations, dark matter halos are
remarkably self-similar. In \citet{NFW1996};\citet{NFW1997}, hereafter
NFW, the authors observe that the spherical radial density profiles
\(\rho(r)\) are well described by a two-parameter law: \begin{equation}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
profiles are \emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases at \(r \sim r_s\) and at large radii, the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example of a NFW halo and a cored dark matter halo.

The total mass of an NFW profile diverges, so halos are commonly
characterized by an overdensity criterion. The virial mass, \(M_{200}\),
is defined as the mass within a region \(r_{200}\) containing a mean
enclosed density 200 times\footnote{200 smells like an arbitrary value.
  From approximate analytic arguments about the mean density of a
  self-gravitating spherical collapse halo in an expanding universe, the
  expected virial overdensity is XXX, rounding up to 200 (REFS).} the
critical density of the universe: \begin{equation}{
M_{200} =(4\pi/3) \ r_{200}^3\ 200\rho_{\rm crit}, \qquad {\rm where} \quad \rho_{\rm crit}(z) = 3H(z) / 8\pi G
}\end{equation} A standard second parameter is the halo concentration,
\(c=r_{200} / r_s\) describing how the characteristic size of the halo
compares to its virial radius. In this case, the scale density is a
function of \(c\) alone,
\(\rho_s = (200/3)\,\rho_{\rm crit} c^3 / [\log(1+c) - c/(1+c)]\)
\citep{NFW1996}. However, we characterize halos by their circular
velocity profiles. The circular velocity,
\(v_{\rm circ}(r) = \sqrt{G M(r) / r}\), reaches a maximum of
\(v_{\rm max}\) at radius \(r_{\rm max} = \alpha\,r_s\) where
\(\alpha\approx2.16258\). \(v_{\rm max}\) is related to both the total
halo mass and the observed line of sight velocity disperion, and
\(r_{\rm max}\) relates to the scale radius.

While two free parameters characterize a NFW halo, \(M_{200}\) and
\(c\), these parameters are not independent. Smaller dark matter halos
often collapse earlier, when the universe was denser. As a result, small
subhalos tend to be more concentrated \citep[e.g.][]{NFW1997}. The
relationship between \(M_{200}\) and c, or the mass-concentration
relation, how more massive halos become less concentrated in a
predictable way \citep[e.g.][\citet{ludlow+2016}]{bullock+2001}. While
there is a general trend in concentration with mass, the concentration
values still have substantial scatter. Other parameters such as the halo
spin or shape or slight changes in the halo density profile add
additional variability to the present-day distribution of halos
\citep[see e.g.][\citet{dutton+maccio2014}, darkEXP,
ect.]{navarro+2010}.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/example_density_profiles.pdf}}
\caption[Example density profiles]{Example stellar and dark matter
density profiles for a Sculptor-like galaxy. The dark matter is more
extended and massive than the star across the entire
galaxy.}\label{fig:nfw_density}
\end{figure}

\subsection{Connecting stars and dark
matter}\label{connecting-stars-and-dark-matter}

Both cosmology and observations find fundamental relationships between
properties of galaxies, particularly for the total mass and luminosity.
A key prediction of \(\Lambda\)CDM is the stellar mass halo mass
relation, describing the amount of stellar mass forms in a (sub)halo of
a given size. In particular, the SMHM relation grows especially steep in
the dwarf galaxy regime---many dwarf galaxies are formed in halos of
similar masses. Fig.~\ref{fig:smhm} shows the stellar mass versus
\(v_{\rm max}\) maximum circular velocity (proxy for halo mass) for
apostle dwarfs from \citet{fattahi+2018}. While there is some scatter,
the range of possible \(v_{\rm max}\) is fairly narrow across \(\sim 5\)
decades in stellar mass. Thus, if we know the initial stellar mass of a
dwarf galaxy, we also know, at some level, its dark matter mass.

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. By being closer to a massive host,
most dwarfs quench earlier, resulting in lower stellar masses
\citep[e.g.][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for Milky Way
satellites {[}@.e.g fattahi+2018{]}. Understanding the effects of tides
on local group dwarf galaxies is essential to determining where and how
these galaxies formed in a cosmological context.

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
\citet{fattahi+2018}. Right: the mass-concentration relation for NFW
halos, but parameterized in terms of \(r_{\rm max}\) and
\(v_{\rm max}\). Together, these plots represent the cosmologically
expected properties of underlying dark matter halos in any dwarf galaxy.
The velocity dispersion directly constrains the mass contained within a
half-light radius, so the underlying halo in \(\Lambda\)CDM is
well-constrained.}\label{fig:smhm}
\end{figure}

@@

Dwarf galaxies have often puzzled astronomers and our understanding of
the universe. While we now generally understand how dwarf galaxies fit
into the universe, this was not always the case. The first discovered
dwarf galaxies in 1938\footnote{technically, the LMC and SMC may be
  classified as dwarf galaxies, but these were likely always known to
  humans at southern latitudes.}, Sculptor and Fornax, heralds an era of
discovery and investigation into these astronomical system.
\citet{shapley1938} present Sculptor and Fornax as a new type of
\emph{stellar system} resembling the Magellanic clouds and globular
clusters, but do not attempt to speculate on the exact nature. The first
direct hints that these systems are galaxies originates from early
spectroscopic work deriving the first velocity dispersions
\citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
\citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
\citet{olszewski1998}.]{aaronson1983}. Based on the physical size and
velocity dispersions, the mass to light ratios of these dwarf galaxies
is far larger than expected for star clusters without dark matter. In
general, these objects have reached a consensus that they have high
velocity dispersion, old spheroidal stellar populations, and low stellar
masses.

Dwarf galaxies span a large range of size and morphology.
Fig.~\ref{fig:galaxy_images} displays several images of dwarf galaxies.
The largest in the local group but not always considered a dwarf galaxy,
the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. Large
LMC-like dwarf galaxies exhibit complex structure such as a bar, young
blue stars, and irregular shapes. On the other hand, the classical and
faint dwarf galaxies are composed mostly of old, faint, and red stars.
These galaxies are typically classified as dwarf spheroidals (dSph)
(although we typically omit the dSph designation and refer to these
galaxies by their constellation alone).

@@

\emph{Classical systems}, the brightest of the local group
(non-Magellanic) dwarf galaxies, are perhaps the systems best studied
and with the most stringently derived properties. Defined\footnote{Note
  that some work like \citet{simon2019} instead define classical dwarfs
  based on luminosity as compared to ultra faint dwarfs. This may
  include galaxies such as Antlia I, Crater II, Canes Venatici I, and
  Sagittarius. We stick with the historical definition for simplicity
  and since the Gaia observations are most numerous for the historical
  classicals.} as the first six discovered dwarf spheroidal galaxies,
(Sculptor, Ursa Minor, Fornax, Draco, Leo I, Leo II, and Carina),
classical systems are typically old, spheroidal, and gas-poor. While
extending 1-2 degrees across the sky (e.g.
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
these dwarfs contain large numbers of bright (giant) stars, allowing
thousands of stars to be observed with deep photometry and spectroscopy
\citep[e.g.][\citet{pace+2020}]{tolstoy+2023}. As a result, the
determination of fundamental properties such as the position, size,
orientation, distance \citep[from 100s of RRL stars,
e.g.][]{tran+2022, garofalo+2025}, proper motions (from \emph{Gaia},
e.g. \citet{battaglia+2022}, \citet{MV2020a}), line-of-sight (LOS)
velocity and dispersion \(\sigma_v\), are all relatively well
constrained today and show excellent convergence among different works.
However, ongoing research continues to redefine our understanding of the
detailed structure of Milky Way satellites, hinting that these objects
may be more complex than at first glance.

@@

For Local group dwarf spheroidal, \emph{Gaia} has allowed for many of
the first proper motion measurements. While the proper motion
uncertainty on a typical dwarf member star is often with large
uncertainties, by combining the proper motions for 100s to 1000s of
stars in \emph{Gaia}, incredibly precise proper motion measurements can
be determined \citep[e.g.][]{MV2020a}, sometimes only limited by
\emph{Gaia}'s systematic error floor. Proper motions have ushered in a
new dynamical era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's (short term) orbit under a given MW
potential.

@@

With the help of \emph{Gaia}, high resolution spectroscopy, high-quality
photometry, and large aperture multi-object spectroscopic instruments,
we can observe dwarf galaxies in unprecedented detail. Today, dwarf
galaxies are used to search for dark matter cores (REFS), test our
understanding of early nucleosynthesis (REFS), the details of galaxy
formation (REFS), and \ldots.

While we now know the basic properties of dwarf galaxies, tremendous
progress has been made in the past 10-20 years with large area deep
photometric surveys, multi-object spectrographs, high resolution
chemistry, and theoretical work, unveiling that dwarf galaxies are each
unique systems with different formation pathways
\citep[e.g.][]{simon2019}.

\citet{battaglia+nipoti2022}

\subsection{Stellar density profiles}\label{stellar-density-profiles}

Given star counts for a local group dwarf, one of the main quantities of
interest is the surface density profile. Long used to characterize and
understand larger galaxies, these profiles can provide hints as to the
formation of dwarf galaxies, determine the physical extent of the
stellar component, and the total stellar mass. Previous works tend to
focus on fitting 3 different surface brightness profiles: an
Exponential, a Plummer, or a King profile.

The exponential profile is perhaps the simplest, as a 1-parameter
profile \begin{equation}{
\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
}\end{equation}

@@

\citet{plummer1911} proposed a 1-parameter solution for a polytropic
potential to fit globular cluster density profiles:

\begin{equation}{
\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2}
}\end{equation}

The \citet{king1962} profile, originating as an empirical fit to
globular clusters, also sometimes reasonably fits dwarf galaxies. This
density profile includes an additional parameter \(R_t\), a truncation
radius where the density sharply drops to 0. \begin{equation}{
\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
}\end{equation}

Finally, the Sérsic profile represents an extension of the exponential.
Typically parameterized in terms of a half-light radius \(R_h\), the
density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\), the
profile's equation is \begin{equation}{
\Sigma_{\rm Sérsic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
}\end{equation} where \(b_n\) is a constant depending on \(n\). Less
commonly used in studies of dwarf galaxies, \citet{munoz+2018} advocate
for using the Sérsic profile since the added flexibility allows for more
profiles to be fit.

While there are not clear theoretical explanations why any profile is
best, the Exponential stellar density profile is commonly assumed as a
single parameter density profile which fits many dwarf galaxies well
\citep[e.g.][\citet{eskridge1988a}, \citet{hodge1991a},
\citet{hodge1991b}, \citet{IH1995}, etc.]{MV2020a} \citet{faber+lin1983}
demonstrated that an Exponential density profile is a reasonable
empirical fit to dwarf galaxies, theorizing that dwarf spheroidal evolve
from exponential disks, maintaining a similar light profile.

in addition to Sérsic, King and other profiles. \citet{lelli+2014}
(example of using exponential.) \citet{wang+2019} for fornax,
\citet{mcconnachie+irwin2006} for Exp, Plummer, King in other galaxies.

However, an exponential profile was often noted to not fit every dwarf
galaxy. Even starting from \citet{aparicio1997}, some people have noted
deviations from this simple empirical rule. Additionally,
\citet{herrmann+hunter+elmegreen2013},
\citet{herrmann+hunter+elmegreen2016} note that at least in a
photometric sample of more distant dwarf disky/Irr/blue compact dwarfs,
that many dwarfs show deviations from exponential profiles.

@@

@@

\begin{table*}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: 1. Muñoz et al. (2018) Sérsic fits, 2. Tran et al. (2022) RR lyrae distance, 3. Alan W. McConnachie and Venn (2020b), 4. Arroyo-Polonio et al. (2024). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012^\circ$ & 1\\
$\delta$ & $-33.7186 \pm 0.0007^\circ$ & ”\\
distance modulus & $19.60 \pm 0.05$ & 2\\
distance & $83.2 \pm 2$ kpc & ”\\
$\mu_{\alpha*}$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & ”\\
LOS velocity & $111.2 \pm 0.3\ {\rm km\,s^{-1}}$ & 4\\
$\sigma_v$ & $9.7\pm0.2\ {\rm km\,s^{-1}}$ & ”\\
$R_h$ & $9.79 \pm 0.04$ arcmin & 1\\
ellipticity & $0.37 \pm 0.01$ & ”\\
position angle & $94\pm1^\circ$ & ”\\
$M_V$ & $-10.82\pm0.14$ & ”\\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018) Sérsic fits, (2) Garofalo et al. (2025) RR lyrae distance, (3) Alan W. McConnachie and Venn (2020a), (4) Pace et al. (2020), average of MMT and Keck results. }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $ 227.2420 \pm 0.0045$˚ & 1\\
$\delta$ & $67.2221 \pm 0.0016$˚ & ”\\
distance modulus & $19.23 \pm 0.11$ & 2\\
distance & $70.1 \pm 3.6$ kpc & ”\\
$\mu_\alpha*$ & $-0.124 \pm 0.004 \pm 0.017$ mas yr$^{-1}$ & 3\\
$\mu_\delta$ & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & ”\\
LOS velocity & $-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ km s$^{-1}$ & 4\\
$\sigma_v$ & $8.6 \pm 0.3$ & ”\\
$R_h$ & $11.62 \pm 0.1$ arcmin & 1\\
ellipticity & $0.55 \pm 0.01$ & ”\\
position angle & $50 \pm 1^\circ$ & ”\\
$M_V$ & $-9.03 \pm 0.05$ & ”\\
\bottomrule
\end{tabular}
\end{table*}
